Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-23 src/2.GenericProgramming.tex
Modified my part of Chapter 2. Expanded OpenMP-like to be more imperative

A programming model is an abstract representation of a computer. It defines a paradigm and a programming language. The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data. Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time. While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.

Parallel programming models add the notion of concurrency to a sequence of instruction of a program. This is commonly achieved either by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations, through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others. Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively. In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at. Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design. We call it the semantic capture.

@@

Lastly, a language defines the syntax of the parallel semantic. Declarative languages add semantic information to the base language through pragma annotations while imperative languages rely on programming interfaces to capture information. Their semantic is defined in the API specification. Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming language types. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} have a similar semantic of describing a parallel loop construct. It is up to the programming model specification to define the execution and memory model in both cases.

@@

\begin{figure}
\begin{Verbatim}[frame=leftline]
# pragma model parallel for
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{Verbatim}
\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
\label{figOMPLike}
\end{figure}

>>>>

A programming model exposes constrained semantics to the programmer which that programmer uses to express intent. Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance portable code. This is commonly achieved by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others. Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen. In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they balance convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly. Questions of what semantic information is needed, what paradigms to present and by what syntactical means span a design space worth taking a closer look at. Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design. We call it the semantic capture.

@@

Lastly, a model must decide between declarative and imperative semantics. Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain. Imperative semantics require the developer to specify the mechanisms by which their intent will be implemented. In an imperative model, developer success is tied to their knowledge of implementation mechanisms. In a declarative model that success is tied to the developer\'s ability to accurately declare their requirements, and the toolchain\'s ability to turn that into an implementation. Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics. While both programming models implement the same programming paradigm (parallel patterns), Figure~\ref{figOMPLike} required the developer to express how to parallelize the loop, while in~\ref{figKokkosLike} the developer just declared their parallel semantics, and relied on the programming model to pick an implementation.

@@

\begin{figure}
\begin{Verbatim}[frame=leftline]
# pragma model parallel for simd vectorize tile unroll fuse
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{Verbatim}
\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
\label{figOMPLike}
\end{figure}
